Investigation of the activities performed by experimental researchers in a software engineering lab using an ethnographic methodology

### 2024-01-03 Response-to-the-reviewers-comments.md

# Responses to Reviewer Comments

We appreciate the editor and reviewers for the comments we have received. We have considered all the comments in the current revision. The improvements are indicated in the manuscript using track changes. The final version of the manuscript can be obtained by replacing the line in the "Article.tex" file:

> \usepackage[draft]{changes}

with:

> \usepackage[final]{changes}

Below, we detail the responses to the comments and the changes made:

**Comment 1** - Improved figure number 2

*Response to Comment 1* - Thank you very much for this observation. Indeed, the purpose of this figure needed to be clarified. We have improved the figure by adding the abstraction of the process that researchers obtained from the sketch made by the experimenters. Additionally, we have added the following explanation in the caption:

> Example of a sketch made by the experimenters. The researchers have used drafts like this to progress in defining the experimental process workflow described in Section 4.1.

**Comment 2** - Proper formatting is required throughout the manuscripts.

*Response to Comment 2* - We have carefully reviewed the manuscript.

- We ensured that the LaTeX template corresponds to the latest version, available at https://digital-library.theiet.org/journals/author-guide. The IET logo now appears in the upper right corner by adjusting the template to the newest version.
- We corrected the way the bibliography is presented in the Reference Section. Cited articles now appear correctly in brackets.
- We corrected the lists so that the bullets appear in the correct place.
- We reduced the size of the bullets.

Since these changes affect the overall format, they are not marked in the manuscript. Otherwise, the entire text would appear as modified.

**Comment 3** - Please provide limitations and future scope in the conclusion section.

*Response to Comment 3* - Threats to validity are described in Section 6 "Threats to validity." However, threats to validity do not represent the limitations of the work. We appreciate the reviewer's comment. In Section 7 "Conclusions," we have included the following text:

> However, this work is subject to various limitations. The first and most important is the empirical basis obtained. Although ethnographic research has been extensive over time and conducted using multiple techniques, in the end, we only have data from one research group. Furthermore, this group's experimenters have self-trained in the field, meaning they have not received specific academic training. The observed peculiarities could be habits acquired over time that other, more recent groups with formal training do not possess.
>
> We also cannot rule out errors in perception on the researchers' part as well. As mentioned earlier, we had yet to learn of experimentation at the beginning of ethnographic research, which may have led us to misinterpret some behaviors of the experimenters. On the other hand, perhaps the researchers' "fresh" perspective has allowed us to identify our findings.

Regarding future work, we have modified the single paragraph that appeared on the manuscript:

> We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore, it is necessary to carry out further studies to generalize the results of the present work. We have completed a survey exploring issues, e.g., terminological diversity, that surfaced during the ethnography. We have also completed a comparative study of the experimental process between the SE and an established engineering discipline. Both works will likely be published shortly. In addition, we plan to conduct more studies (probably not ethnographic because they require too much time) on this topic soon.

replacing it with a more explicit and detailed description of future research plans:

> To overcome the limitations of the present investigation, it is necessary to carry out further studies to generalize the results of the current work. Our plans are as follows:
> - Verify the existence of terminological diversity, one of the first aspects that has emerged in the research and is relatively easy to check. We have already started this work. We surveyed SE experimenters from multiple research groups. In the survey, we asked them about the meaning and usage of experimentation-related terms. Preliminary results confirm the existence of terminological diversity.
>- After completing the ethnographic research, we initiated a comparative study of the experimental process between SE and an established engineering discipline (specifically, biotechnology). Preliminary results indicate the absence of terminological diversity and a greater formalization of the experimental process within specific research areas but not at a general level. This could be the key to terminological diversity in SE: sub-areas of SE generally have not developed specific terminology and protocols (with [51] being a notable exception), which could lead to the need for ad-hoc concepts and processes.
>- Both works will likely be published shortly. In addition, we plan to conduct more studies in the future on this topic soon. We still need to decide on the objects under study and the methodologies used since ethnography is very demanding. A possibility to study the impact of specialization in the sciences is to analyze experimental repositories. This represents an indirect but efficient approach in terms of time and effort. Another alternative is conducting qualitative studies using focus groups composed of SE experimental researchers.

**Comment 4** - Please clearly define the objective of manuscripts in the introduction section.

*Response to Comment 4* - Thank you very much. Indeed, the definition of the research objective was a single and short line. We have re-written the research objective more clearly:

> This paper aims to investigate how experimental researchers conduct experiments in practice. We intend to determine how experimental researchers plan, execute, analyze, and report their investigations and what concepts, protocols, and processes they use.

**Comment 5** - Please add some recent references.

*Response to Comment 5* - We have included the following recent references:

> Traini, L.: 'Exploring performance assurance practices and challenges in agile software development: an ethnographic study', Empirical Software Engineering, 2022, 27, (3), pp. 74.

> Azarova, M., Hazoglou, M., Aronoff.Spencer, E.: 'Just slack it: A study of multidisciplinary teamwork based on ethnography and data from online collaborative software', New Media & Society, 2022, 24, (6), pp. 1435–1458.

>S. Yadav, M. S. I. Sudman, P. Kumar Dubey, R. Vijava Srinivas, R. Srisainath and V. Chithra Devi, "Development of an GA-RBF based Model for Penetration of Electric Vehicles and its Projections," 2023 International Conference on Self Sustainable Artificial Intelligence Systems (ICSSAS), Erode, India, 2023, pp. 1-6, doi: 10.1109/ICSSAS57918.2023.10331883.

### 2024-01-03 Article.tex
Last changes
Some minor changes
Cambios necesarios para contestar a la primera revision de IET

%%%%%%Begin MyPackages%%
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{framed}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{color}
\usepackage{tabulary}
\usepackage{enumitem}
	\setlist[itemize]{leftmargin=5.5mm}

%\usepackage{hyperref}
%\usepackage{url}
\usepackage[hyphens]{url}
\usepackage[hidelinks]{hyperref}
\hypersetup{breaklinks=true}
\urlstyle{same}
%\usepackage{cite}
\usepackage{soul}
\usepackage{rotating}
\usepackage{lscape}
\usepackage{caption}
\usepackage[title]{appendix}
\usepackage{ulem}
\usepackage{placeins}
\usepackage{balance}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}
%%%%%%end MyPackages%%%%

@@

\title{An ethnographic study of an empirical software engineering lab}

@@

\begin{comment}

\section{Insert A head here}
This demo file is intended to serve as a ``starter file''
for IMAIAI journal papers produced under \LaTeX\ using
imaiai.cls v1.5e.

\subsection{Insert B head here}
Subsection text here.

\subsubsection{Insert C head here}
Subsubsection text here.

\section{Equations}

Sample equations.

%%% Numbered equation
\begin{align}\label{1.1}
\begin{split}
\frac{\partial u(t,x)}{\partial t} &= Au(t,x) \left(1-\frac{u(t,x)}{K}\right)-B\frac{u(t-\tau,x) w(t,x)}{1+Eu(t-\tau,x)},\\
\frac{\partial w(t,x)}{\partial t} &=\delta \frac{\partial^2w(t,x)}{\partial x^2}-Cw(t,x)+D\frac{u(t-\tau,x)w(t,x)}{1+Eu(t-\tau,x)},
\end{split}
\end{align}

\begin{align}\label{1.2}
\begin{split}
\frac{dU}{dt} &=\alpha U(t)(\gamma -U(t))-\frac{U(t-\tau)W(t)}{1+U(t-\tau)},\\
\frac{dW}{dt} &=-W(t)+\beta\frac{U(t-\tau)W(t)}{1+U(t-\tau)}.
\end{split}
\end{align}

%%%% Unnumbered equation
\[
\frac{\partial(F_1,F_2)}{\partial(c,\omega)}_{(c_0,\omega_0)} = \left|
\begin{array}{ll}
\frac{\partial F_1}{\partial c} &\frac{\partial F_1}{\partial \omega} \\\noalign{\vskip3pt}
\frac{\partial F_2}{\partial c}&\frac{\partial F_2}{\partial \omega}
\end{array}\right|_{(c_0,\omega_0)}
\]

\section{Enunciations}
%%%% Most of the enunciations like theorem, lemma, corollary, proposition, defintion,
%%%% condition, example, conjecture etc. are defined in the class file.

%%%% If the author wants to add or modify the enunciation style
%%%% they can define in the preamble as shown below.

%%%% \newtheoremstyle{theorem}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
%%%% \theoremstyle{theorem}
%%%% \newtheorem{theorem}{\sc Theorem}[section]

%%%%\newtheoremstyle{corollary}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
%%%%\theoremstyle{corollary}
%%%%\newtheorem{corollary}{\sc Corollary}[section]

%%%%\newtheoremstyle{definition}{6pt}{6pt}{\rm}{}{\sffamily}{ }{ }{}
%%%%\theoremstyle{definition}
%%%%\newtheorem{definition}[theorem]{\sc Definition}
%%%%
%%%%\newtheorem{exercise}[theorem]{Exercise}

\begin{theorem}\label{T0.1}
Assume that $\alpha>0, \gamma>1, \beta>\frac{\gamma+1}{\gamma-1}$.
Then there exists a small $\tau_1>0$, such that for $\tau\in
[0,\tau_1)$, if $c$ crosses $c(\tau)$ from the direction of
to  a small amplitude periodic traveling wave solution of
(2.1), and the period of $(\check{u}^p(s),\check{w}^p(s))$ is
\[
\check{T}(c)=c\cdot \left[\frac{2\pi}{\omega(\tau)}+O(c-c(\tau))\right].
\]
\end{theorem}

\section{Figures \& Tables}

The output for figure is:

\begin{figure}[!h]
%\centering\includegraphics[width=2.5in]{figurename.eps}
%%%call your figure name in the place "figurename.eps"
\caption{Insert figure caption here}
\label{fig_sim}
\end{figure}

 An example of a double column floating figure using two subfigures.
 (The subfig.sty package must be loaded for this to work.)
 The subfigure \verb+\label+ commands are set within each subfloat command, the
 \verb+\label+ for the overall figure must come after \verb+\caption+.
 \verb+\hfil+ must be used as a separator to get equal spacing.
 The subfigure.sty package works much the same way, except \verb+\subfigure+ is
 used instead of \verb+\subfloat+.

%\begin{figure*}[!h]
%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{figurename.eps}%
%\label{fig_first_case}}
%\hfil
%\subfloat[Case II]{\includegraphics[width=2.5in]{figurename.eps}%
%\label{fig_second_case}}}
%\caption{Simulation results}
%\label{fig_sim}
%\end{figure*}

\vskip2pc

\noindent The output for table is:

\begin{table}[!h]
\caption{An Example of a Table}%%%Table caption goes here
\label{table_example}
\centering
\begin{tabular}{|c||c|}%%%The number of columns has to be defined here
\hline
One & Two\\ %%%% Table body
\hline
Three & Four\\%%%% Table body
\hline
\end{tabular}
\end{table}%%%End of the table

\section{Conclusion}
The conclusion text goes here.
\end{comment}

\section*{Acknowledgment}
This work was supported by the project PGC2018-097265-B-I00 funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII). This work's preprint has been published \cite{Fonseca-2023-experimentation-proccess-SE-unpublished}.

@@

\Urlmuskip=0mu plus 1mu\relax
\bibliographystyle{iet}
\bibliography{bibliography}
%
% once the .bbl file has been generated then place the text in your article.

% To get the numbered reference style the author should use [numbib]
%as an option in the document class.  For example: \documentclass[numbib]{imaiai}
\begin{comment}

\begin{thebibliography}{9}
\bibitem{1}
Vicsek T., Czir\'{o}k A., Ben-Jacob E., Cohen I.: `Novel type of phase transition in a system of self-driven particles', \textit{Phys. Rev. Lett.}, 1995, \textbf{75}, pp.~1226--1229

\bibitem{2}
Jadbabaie A., Lin J., Morse A.S.: `Coordination of groups of
mobile autonomous agents using nearest neighbor rules', \textit{IEEE
Trans. Autom. Control}, 2003, \textbf{48}, pp.~988--1001

\bibitem{3}
Olfati-Saber R., Murray R.M.: `Consensus problems in networks
of agents with switching topology and time-delays', \textit{IEEE Trans.
Autom. Control}, 2004, \textbf{49}, pp.~1520--1533
\end{thebibliography}
\end{comment}

>>>>

%%%%%%Begin MyPackages%%
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{framed}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{color}
\usepackage{tabulary}
\usepackage{enumitem}
	\setlist[itemize]{leftmargin=5.5mm}

%\usepackage{hyperref}
%\usepackage{url}
\usepackage[hyphens]{url}
\usepackage[hidelinks]{hyperref}
\hypersetup{breaklinks=true}
\urlstyle{same}

\usepackage{natbib}
\setcitestyle{square}

\usepackage{soul}
\usepackage{rotating}
\usepackage{lscape}
\usepackage{caption}
\usepackage[title]{appendix}
\usepackage{ulem}
\usepackage{placeins}
\usepackage{balance}
\usepackage[autostyle, english = american]{csquotes}
\MakeOuterQuote{"}

\usepackage[draft]{changes}

%%%%%%end MyPackages%%%%

%% Para arreglar las viñetas

\setitemize{label={\raisebox{0pt}{\normalsize\textbullet}}, labelindent=0em,labelsep=1em,leftmargin=*}

%% Hay otros cambios en el .cls

@@

\supertitle{Submission Template for IET Research Journal Papers}

\title{Investigation of the activities performed by experimental researchers in a software engineering lab using an ethnographic methodology}

@@

\section*{Acknowledgment}
This work was supported by the project \replaced{PGC2018-097265-B-I00}{PID2022-137846NB-I00} funded by the Spanish Ministry of Science and Innovation and co-financed by FEDER, and the Universidad de las Fuerzas Armadas  - ESPE's Grupo de Investigación en Modelos de Producción de Software e Industria Inteligente (GrIMPSoftII). This work's preprint has been published \cite{Fonseca-2023-experimentation-proccess-SE-unpublished}.

@@

\Urlmuskip=0mu plus 1mu\relax
\bibliographystyle{iet}
\bibliography{bibliography}

### 2024-01-03 tex/conclusion.tex
Cambios necesarios para contestar a la primera revision de IET

We believe that the "lab mythology" is worth studying. If the differential characteristics of the theoretical vs. actual experiment process are specific to the research group, the inquiry will end. However, it is also possible that widespread patterns of behavior show up. Bringing such knowledge to the foreground could contribute to improving experimental practice.

We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore, it is necessary to carry out further studies to generalize the results of the present work. We have completed a survey exploring issues, e.g., terminological diversity, that surfaced during the ethnography. We have also completed a comparative study of the experimental process between the SE and an established engineering discipline. Both works will likely be published shortly. In addition, we plan to conduct more studies (probably not ethnographic because they require too much time) on this topic soon.

>>>>

\added{However, this work is subject to various limitations. The first and most important is the empirical basis obtained. Although ethnographic research has been extensive over time and conducted using multiple techniques, in the end, we only have data from one research group. Furthermore, this group's experimenters have self-trained in the field, meaning they have not received specific academic training. The observed peculiarities could be habits acquired over time that other, more recent groups with formal training do not possess.}

\added{We also cannot rule out errors in perception on the researchers' part as well. As mentioned earlier, we had yet to learn of experimentation at the beginning of ethnographic research, which may have led us to misinterpret some behaviors of the experimenters. On the other hand, perhaps the researchers' "fresh" perspective has allowed us to identify our findings.}

\deleted{We believe that the ''lab mythology'' is worth studying. If the differential characteristics of the theoretical vs. actual experiment process are specific to the research group, the inquiry will simply end. However, it is also possible that widespread patterns of behavior show up. Bringing such knowledge to the foreground could contribute to improving experimental practice.}

\deleted{We are aware that the findings of this study correspond to the particular perspective of an experimental research group, albeit representative of the ESE community. Therefore,}
\added{To overcome the limitations of the present investigation,} it is necessary to carry out further studies to generalize the results of the current work. \added{Our plans are as follows:}
\begin{itemize}

\item \added{Verify the existence of terminological diversity, one of the first aspects that has emerged in the research and is relatively easy to check. We have already started this work. We surveyed SE experimenters from multiple research groups. In the survey, we asked them about the meaning and usage of experimentation-related terms. Preliminary results confirm the existence of terminological diversity.}

\item \added{After completing the ethnographic research, we initiated a comparative study of the experimental process between SE and an established engineering discipline (specifically, biotechnology). Preliminary results indicate the absence of terminological diversity and a greater formalization of the experimental process within specific research areas but not at a general level. This could be the key to terminological diversity in SE: sub-areas of SE generally have not developed specific terminology and protocols (with \cite{Arcuri-2014-randomized} being a notable exception), which could lead to the need for ad-hoc concepts and processes.}

\item \replaced{Both works will likely be published shortly. In addition, we plan}{We plan} to conduct more studies \replaced{in the future}{(probably not ethnographic because they require too much time)} on this topic soon. \added{We still need to decide on the objects under study and the methodologies used since ethnography is very demanding. A possibility to study the impact of specialization in the sciences is to analyze experimental repositories. This represents an indirect but efficient approach in terms of time and effort. Another alternative is conducting qualitative studies using focus groups composed of SE experimental researchers.}

\end{itemize}

\replaced{In short, w}{W}e believe that the "lab mythology" is worth studying. If the differential characteristics of the theoretical vs. actual experiment process \replaced{were}{are} specific to the research group, the inquiry \replaced{would}{will } simply end. However, it is also possible that widespread patterns of behavior show up. Bringing such knowledge to the foreground could contribute to improving experimental practice.

### 2024-01-03 tex/introduction.tex
Cambios necesarios para contestar a la primera revision de IET

This paper aims to determine \textbf{how experimental researchers conduct experiments in practice}. We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities they perform daily. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different subareas or families of experiments have about the overall process.

>>>>

This paper aims to \replaced{investigate}{determine} \textbf{how experimental researchers conduct experiments in practice}. \added{We intend to find out how experimental researchers plan, execute, analyze, and report their investigations and what concepts, protocols, and processes they use.}

We applied an ethnographic method \cite{Sharp-2016-Ethnographic-Studies-ESE} within an experienced experimental research group. We observed the researchers in their lab for two years. We created conceptual and process models to represent the concepts they use and the activities they perform daily. These models fit the community's procedures and terminology at a high level. Still, the models show particularities at lower levels: (1) The number and diversity of activities, (2) the existence of different roles, (3) the granularity of the concepts used by the roles, and (4) the viewpoints that different subareas or families of experiments have about the overall process.

### 2024-01-03 tex/research-execution.tex
Cambios necesarios para contestar a la primera revision de IET

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Process-Approximation-Draft}
	\caption{example of a sketch made by experimenters.}
	\label{fig-proceso-exp-boseto}
\end{figure}

>>>>

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Process-Approximation-Draft}
	\caption{Example of a sketch made by the experimenters. \added{The researchers have used drafts like this to progress in the definition of the experimental process workflow described in Section~\ref{sec-results:process-workflow}.}}
	\label{fig-proceso-exp-boseto}
	\highlight{This new figure replaces the previous Figure 2, which was included}
	\highlight{in the first version of the manuscript.}
\end{figure}

### 2024-01-03 tex/research-method.tex
Cambios necesarios para contestar a la primera revision de IET

\section{Research Method}\label{sec-research-method}
Ethnography is a \textit{participant observation} method. It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}. Ethnographic researchers adopt "the ESE researchers' point of view" to better understand the reason for their actions \cite[pp. 55-56]{Outhwaite-2007-sage}. Thus, the biases and misperceptions that ESE researchers might have about themselves can be detected. Ethnography also captures the rich context that surrounds ESE researchers.

Ethnography identifies details that are ignored by or in other methods. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted to study a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.

>>>>

\section{Research Method}\label{sec-research-method}

Ethnography is a \textit{participant observation} method \added{used with some frequency in SE, e.g., \cite{traini2022exploring, azarova2022just}}. It does not interfere (beyond the ethnographic researchers' presence) with the observed ESE researchers' "life" \cite{Sharp-2016-Ethnographic-Studies-ESE}. Ethnographic researchers adopt "the ESE researchers' point of view" to better understand the reason for their actions \cite[pp. 55-56]{Outhwaite-2007-sage}. Thus, the biases and misperceptions that ESE researchers might have about themselves can be detected. Ethnography also captures the rich context that surrounds ESE researchers.

Ethnography identifies details that are ignored by \deleted{or in} other method\replaced{ologies, including Artificial Intelligence (AI) \cite{yadav2023development}}{s}. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted to study a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.
